---
title: "WEB - Assotest"
date: 'Page loaded: `r format(as.POSIXct(format(Sys.time())), "%b %d, %Y")`'
output:
  flexdashboard::flex_dashboard:
    navbar:
      - { title: "About", href: "#about", align: right, icon: glyphicon-info-sign}
    orientation: rows
    vertical_layout: fill
    social: menu
    theme:
      version: 4
      base_font:
        google: Prompt
      heading_font:
        google: Sen
runtime: shiny
---


<style type="text/css">
  body {
  font-size: 16pt;
}

h1,h2,h3,h4,h5,h6 {
  font-size: 24pt;
}

.navbar-brand {
  color:white;
  font-size: 32px;
}

div.navbar-header {
  font-size: 28pt;
}

div.chart-title {
  font-size: 28pt;
}

/* Center the page-loaded date in the navbar */
.navbar-author {
  position: absolute;
  left: 50%;
  transform: translateX(-50%);
  color: rgba(255,255,255,0.8);
  line-height: 50px;
  white-space: nowrap;
  pointer-events: none;
}

/* Hide the About page tab from the left-side tab navigation only */
.navbar-nav li:has(a[data-toggle="tab"][href="#about"]) {
  display: none !important;
}

/* Left-align rhandsontable row header labels */
.handsontable tbody th {
  text-align: left !important;
}
</style>



```{r setup, include=FALSE}
library("flexdashboard")
library("rhandsontable")
library("shinyscreenshot")

# Default genotype count table — rows: Cases/Controls, cols: WT/Het/Hom
df <- data.frame(
  WT  = c(10L, 10L),
  Het = c(20L, 20L),
  Hom = c(30L, 30L)
)
rownames(df) <- c("Cases", "Controls")


# --------------------------------------------------------------------
# SNPHWE(obs_hets, obs_hom1, obs_hom2)
# Exact test of Hardy-Weinberg Equilibrium.
# Reference: Wigginton JE, Cutler DJ, Abecasis GR (2005). A Note on
# Exact Tests of Hardy-Weinberg Equilibrium. Am J Hum Genet 76:887-893.
#
# obs_hom1, obs_hom2: homozygote counts (order does not matter;
#   the algorithm identifies the rare homozygote internally via min/max).
# obs_hets: heterozygote count.
# Returns the two-sided p-value, or -1.0 on invalid (negative) input.
#
# Algorithm: enumerates the conditional distribution of heterozygote
# counts given fixed marginals, computing probabilities by recurrence
# outward from the mode to avoid overflow.
# --------------------------------------------------------------------
SNPHWE <- function(obs_hets, obs_hom1, obs_hom2) {
  if (obs_hom1 < 0 || obs_hom2 < 0 || obs_hets < 0)
    return(-1.0)

  N       <- obs_hom1 + obs_hom2 + obs_hets
  obs_homr <- min(obs_hom1, obs_hom2)
  obs_homc <- max(obs_hom1, obs_hom2)
  rare    <- obs_homr * 2 + obs_hets

  probs <- rep(0, 1 + rare)

  mid <- floor(rare * (2 * N - rare) / (2 * N))
  if ((mid %% 2) != (rare %% 2)) mid <- mid + 1

  probs[mid + 1] <- 1.0
  mysum <- 1.0

  # Probabilities from midpoint downward
  curr_hets <- mid
  curr_homr <- (rare - mid) / 2
  curr_homc <- N - curr_hets - curr_homr

  while (curr_hets >= 2) {
    probs[curr_hets - 1] <- probs[curr_hets + 1] * curr_hets *
      (curr_hets - 1.0) / (4.0 * (curr_homr + 1.0) * (curr_homc + 1.0))
    mysum     <- mysum + probs[curr_hets - 1]
    curr_hets <- curr_hets - 2
    curr_homr <- curr_homr + 1
    curr_homc <- curr_homc + 1
  }

  # Probabilities from midpoint upward
  curr_hets <- mid
  curr_homr <- (rare - mid) / 2
  curr_homc <- N - curr_hets - curr_homr

  while (curr_hets <= rare - 2) {
    probs[curr_hets + 3] <- probs[curr_hets + 1] * 4.0 * curr_homr *
      curr_homc / ((curr_hets + 2.0) * (curr_hets + 1.0))
    mysum     <- mysum + probs[curr_hets + 3]
    curr_hets <- curr_hets + 2
    curr_homr <- curr_homr - 1
    curr_homc <- curr_homc - 1
  }

  target <- probs[obs_hets + 1]
  p <- min(1.0, sum(probs[probs <= target]) / mysum)
  return(p)
}


# --------------------------------------------------------------------
# check_data_validity(ca, co)
# ca, co: integer vectors of length 3 (WT, Het, Hom counts).
# Returns list(valid = TRUE/FALSE, msg = character).
# Rejects tables where all cases or all controls are zero, or where
# fewer than 2 genotype groups have any observations (models cannot
# be identified with only one non-empty column).
# --------------------------------------------------------------------
check_data_validity <- function(ca, co) {
  if (sum(ca + co == 0) > 1) return(list(valid = FALSE,
    msg = "At least two genotype groups must have observations."))
  if (sum(ca) == 0) return(list(valid = FALSE,
    msg = "There are no cases. Please enter case counts."))
  if (sum(co) == 0) return(list(valid = FALSE,
    msg = "There are no controls. Please enter control counts."))
  return(list(valid = TRUE, msg = ""))
}


# --------------------------------------------------------------------
# hwe_pvalues(ca, co, method)
# ca, co: integer vectors of length 3. method: "chisq" or "exact".
# Returns a list with test statistics (t.ca, t.co, t.2) and p-values
# (p.ca, p.co, p.2) for cases, controls, and both combined.
#
# For method = "exact": individual statistics are NA (SNPHWE returns
# only a p-value); the combined p-value uses Fisher's method —
# -2*(log(p.ca) + log(p.co)) ~ chi-sq(4).
# For method = "chisq": combined statistic is the sum of the two
# chi-square values ~ chi-sq(2).
# --------------------------------------------------------------------
hwe_pvalues <- function(ca, co, method = "chisq") {

  # Chi-square HWE test for a single group x = c(WT, Het, Hom)
  hwe_chisq <- function(x) {
    f   <- (2 * x[1] + x[2]) / (2 * sum(x))   # reference allele frequency
    e   <- c(f^2, 2 * f * (1 - f), (1 - f)^2) * sum(x)
    tst <- sum((x - e)^2 / e)
    p   <- 1 - pchisq(tst, 1)
    list(stat = tst, p = p)
  }

  if (method == "exact") {
    p.ca  <- SNPHWE(ca[2], ca[1], ca[3])
    p.co  <- SNPHWE(co[2], co[1], co[3])
    # Combined: sum of -2*log(p) ~ chi-sq(4) (Fisher's method)
    t.2   <- -2 * (log(p.ca) + log(p.co))
    p.2   <- 1 - pchisq(t.2, 4)
    t.ca  <- NA
    t.co  <- NA
  } else {
    r.ca  <- hwe_chisq(ca)
    r.co  <- hwe_chisq(co)
    t.ca  <- r.ca$stat;  p.ca <- r.ca$p
    t.co  <- r.co$stat;  p.co <- r.co$p
    t.2   <- t.ca + t.co
    p.2   <- 1 - pchisq(t.2, 2)
  }

  list(t.ca = t.ca, p.ca = p.ca,
       t.co = t.co, p.co = p.co,
       t.2  = t.2,  p.2  = p.2)
}


# --------------------------------------------------------------------
# ci.mat(alpha)
# Returns a 2x3 matrix for simultaneous Wald CI computation.
# Post-multiply a [estimate, se] matrix by ci.mat() to obtain
# [estimate, lower, upper] in one step — e.g.:
#   exp(summary(fit)$coef[, 1:2] %*% ci.mat())
# gives the OR with 95% CIs directly from log-OR and its SE.
# --------------------------------------------------------------------
ci.mat <- function(alpha = 0.05) {
  ciM <- rbind(c(1, 1, 1), qnorm(1 - alpha / 2) * c(0, -1, 1))
  colnames(ciM) <- c("Estimate",
    paste0(formatC(100 *    alpha / 2,  format = "f", digits = 1), "%"),
    paste0(formatC(100 * (1 - alpha / 2), format = "f", digits = 1), "%"))
  ciM
}


# --------------------------------------------------------------------
# gtm(ca, co, flev, fnam, main)
# ca, co: integer vectors of length 3 (WT, Het, Hom counts).
# flev: genotype level labels. fnam/main: plot title.
#
# Fits five nested logistic regression models via glm(..., binomial):
#   nul: intercept only (no association)
#   gen: factor(WT/Het/Hom) — genotype model, 2 df
#   dom: factor(WT vs Het+Hom) — dominant model, 1 df
#   cod: linear score 1/2/3 — co-dominant/additive model, 1 df
#   rec: factor(WT+Het vs Hom) — recessive model, 1 df
#
# Draws a model-comparison diagram using base graphics. Lines connect
# nested models; lwd=8 when the simpler model cannot be rejected at
# p >= 0.05 (compatible), lwd=2 otherwise (rejected).
#
# When any cell in a contrast is zero, Fisher's exact test replaces
# the logistic-regression OR and CI for that contrast.
#
# Returns a summary data frame invisibly (used by the CSV download).
# --------------------------------------------------------------------
gtm <- function(ca, co,
                flev = c("WT", "Het", "Hom"),
                fnam = " ",
                main = fnam) {

  rlev <- c(rep(paste(flev[1:2], collapse = "/"), 2), flev[3])
  dlev <- c(flev[1], rep(paste(flev[2:3], collapse = "/"), 2))
  f.g  <- factor(flev)
  f.d  <- factor(dlev, levels = unique(dlev))
  v.c  <- 1:3
  f.r  <- factor(rlev, levels = unique(rlev))

  par(mar = c(0, 0, 2, 0), cex = 1.6)
  plot(NA, type = "n", axes = FALSE, xlab = "", ylab = "",
       xlim = c(-10, 100), ylim = c(0, 100), main = main)

  # Guard: warn if data are marginal but still plottable
  if (sum(c(ca, co) == 0) > 3)
    text(-10, 80, "Warning: many zero cells — interpret with caution.",
         font = 2, adj = 0)

  # Fit models
  nul <- glm(cbind(ca, co) ~ 1,   family = binomial)
  gen <- glm(cbind(ca, co) ~ f.g, family = binomial)
  dom <- glm(cbind(ca, co) ~ f.d, family = binomial)
  cod <- glm(cbind(ca, co) ~ v.c, family = binomial)
  rec <- glm(cbind(ca, co) ~ f.r, family = binomial)

  # Likelihood-ratio tests for all nested pairs
  t.dn  <- anova(nul, dom, test = "Chisq")
  t.cn  <- anova(nul, cod, test = "Chisq")
  t.rn  <- anova(nul, rec, test = "Chisq")
  t.gd  <- anova(dom, gen, test = "Chisq")
  t.gc  <- anova(cod, gen, test = "Chisq")
  t.gr  <- anova(rec, gen, test = "Chisq")
  t.gn  <- anova(nul, gen, test = "Chisq")

  extract <- function(a) list(chisq = abs(a[2, 4]), p = a[2, 5])
  r.dn <- extract(t.dn); r.cn <- extract(t.cn); r.rn <- extract(t.rn)
  r.gd <- extract(t.gd); r.gc <- extract(t.gc); r.gr <- extract(t.gr)
  r.gn <- extract(t.gn)

  # Odds ratio estimates (log-OR ± SE → OR with CI via ci.mat)
  OR.g <- exp(summary(gen)$coef[-1, 1:2, drop = FALSE] %*% ci.mat())
  OR.d <- exp(summary(dom)$coef[-1, 1:2, drop = FALSE] %*% ci.mat())
  OR.c <- exp(summary(cod)$coef[-1, 1:2, drop = FALSE] %*% ci.mat())
  OR.r <- exp(summary(rec)$coef[-1, 1:2, drop = FALSE] %*% ci.mat())

  # Pad OR matrices in case of singularities (complete separation)
  if (nrow(OR.g) < 2) OR.g <- rbind(OR.g, rep(NA, 3))
  if (nrow(OR.d) == 0) OR.d <- matrix(NA, 1, 3)
  if (nrow(OR.r) == 0) OR.r <- matrix(NA, 1, 3)

  # Fall back to Fisher's exact test when zeros are present.
  # Each block checks the 2x2 table for the relevant contrast;
  # if any cell is zero the logistic OR/CI is replaced by Fisher's.
  gt1 <- rbind(co, ca)[, -3]   # WT vs Het
  gt2 <- rbind(co, ca)[, -2]   # WT vs Hom
  if (any(gt1 == 0)) {
    ft <- fisher.test(gt1)
    OR.g[1, ] <- c(ft[["estimate"]], ft[["conf.int"]])
  }
  if (any(gt2 == 0)) {
    ft <- fisher.test(gt2)
    OR.g[2, ] <- c(ft[["estimate"]], ft[["conf.int"]])
  }
  dot <- rbind(co, ca) %*% cbind(c(1, 0, 0), c(0, 1, 1))  # WT vs Het+Hom
  if (any(dot == 0)) {
    ft <- fisher.test(dot)
    OR.d[1, ] <- c(ft[["estimate"]], ft[["conf.int"]])
  }
  if (sum(wh <- (ca + co > 0)) < 3) {
    cot <- rbind(co, ca)[, wh]
    if (any(cot == 0)) {
      ft <- fisher.test(cot)
      OR.c[1, ] <- c(ft[["estimate"]], ft[["conf.int"]])
      # When the Het column is missing, the additive OR is per 2 copies;
      # take the square root to recover the per-allele OR.
      if (!wh[2]) OR.c <- sqrt(OR.c)
    }
  }
  rct <- rbind(co, ca) %*% cbind(c(1, 1, 0), c(0, 0, 1))  # WT+Het vs Hom
  if (any(rct == 0)) {
    ft <- fisher.test(rct)
    OR.r[1, ] <- c(ft[["estimate"]], ft[["conf.int"]])
  }

  # Genotype model — top centre
  text(50, 98, "Genotype model", font = 2)
  text(50, 95, paste("OR(", flev[2], "vs.", flev[1], "):",
                     formatC(OR.g[1, 1], format = "f", digits = 2),
                     "(", formatC(OR.g[1, 2], format = "f", digits = 2),
                     ";", formatC(OR.g[1, 3], format = "f", digits = 2), ")"))
  text(50, 92, paste("OR(", flev[3], "vs.", flev[1], "):",
                     formatC(OR.g[2, 1], format = "f", digits = 2),
                     "(", formatC(OR.g[2, 2], format = "f", digits = 2),
                     ";", formatC(OR.g[2, 3], format = "f", digits = 2), ")"))

  # Dominant model — left
  text(0, 53, "Dominant", adj = c(0, 0.5), font = 2)
  text(0, 50, paste("OR(", dlev[2], "vs.", dlev[1], "):"), adj = c(0, 0.5))
  text(0, 47, paste(formatC(OR.d[1, 1], format = "f", digits = 2),
                    "(", formatC(OR.d[1, 2], format = "f", digits = 2),
                    ";", formatC(OR.d[1, 3], format = "f", digits = 2), ")"),
       adj = c(0, 0.5))

  # Co-dominant model — centre
  text(50, 53, "Co-dominant", adj = c(0.5, 0.5), font = 2)
  text(50, 50, paste("OR(", flev[3], "vs.", flev[2], ")=",
                     "OR(", flev[2], "vs.", flev[1], "):"), adj = c(0.5, 0.5))
  text(50, 47, paste(formatC(OR.c[1, 1], format = "f", digits = 2),
                     "(", formatC(OR.c[1, 2], format = "f", digits = 2),
                     ";", formatC(OR.c[1, 3], format = "f", digits = 2), ")"),
       adj = c(0.5, 0.5))

  # Recessive model — right
  text(100, 53, "Recessive", adj = c(1, 0.5), font = 2)
  text(100, 50, paste("OR(", rlev[3], "vs.", rlev[1], "):"), adj = c(1, 0.5))
  text(100, 47, paste(formatC(OR.r[1, 1], format = "f", digits = 2),
                      "(", formatC(OR.r[1, 2], format = "f", digits = 2),
                      ";", formatC(OR.r[1, 3], format = "f", digits = 2), ")"),
       adj = c(1, 0.5))

  # Null model — bottom centre
  text(50, 8, "Null model", adj = c(0.5, 0.5), font = 2)
  text(50, 5, "OR = 1", adj = c(0.5, 0.5))
  text(50, 2, substitute(chi^2 * rest,
    list(rest = paste("(2)=", formatC(r.gn$chisq, format = "f", digits = 3),
                      ", p=", formatC(r.gn$p,     format = "f", digits = 3)))),
    adj = c(0.5, 0.5))

  # Connecting lines: lwd=8 when p >= 0.05 (simpler model not rejected),
  # lwd=2 when rejected. Thick lines indicate compatible simplifications.
  lwd_for <- function(p) if (!is.na(p) && p >= 0.05) 8 else 2

  lines(c(40, 10), c(90, 55), lwd = lwd_for(r.gd$p))  # Genotype  -> Dominant
  lines(c(10, 40), c(45, 10), lwd = lwd_for(r.dn$p))  # Dominant  -> Null
  lines(c(50, 50), c(10, 45), lwd = lwd_for(r.cn$p))  # Null      -> Co-dominant
  lines(c(50, 50), c(55, 90), lwd = lwd_for(r.gc$p))  # Co-dominant -> Genotype
  lines(c(60, 90), c(90, 55), lwd = lwd_for(r.gr$p))  # Genotype  -> Recessive
  lines(c(90, 60), c(45, 10), lwd = lwd_for(r.rn$p))  # Recessive -> Null

  rect(0, 25, 100, 30, col = "white", border = "white")
  rect(0, 70, 100, 75, col = "white", border = "white")

  # Test statistics along connecting lines
  # Order matches line segments: gd, dn, cn, gc, gr, rn
  results_list <- list(r.gd, r.dn, r.cn, r.gc, r.gr, r.rn)
  xt <- c(30, 30, 50, 50, 70, 70)
  yt <- c(72.5, 27.5, 27.5, 72.5, 72.5, 27.5)
  at <- c(0.9, 0.9, 0.5, 0.5, 0.1, 0.1)

  for (i in 1:6)
    text(xt[i], yt[i],
         substitute(chi^2 == rest,
           list(rest = paste(formatC(results_list[[i]]$chisq, format = "f", digits = 3),
                             ", p=",
                             formatC(results_list[[i]]$p,     format = "f", digits = 3)))),
         adj = c(at[i], 0.5))

  # Return a tidy summary data frame for download
  invisible(
    data.frame(
      Model     = c("Dominant", "Co-dominant", "Recessive", "Genotype"),
      OR        = c(OR.d[1,1], OR.c[1,1], OR.r[1,1], NA),
      CI_lower  = c(OR.d[1,2], OR.c[1,2], OR.r[1,2], NA),
      CI_upper  = c(OR.d[1,3], OR.c[1,3], OR.r[1,3], NA),
      Chisq     = c(r.dn$chisq, r.cn$chisq, r.rn$chisq, r.gn$chisq),
      p_value   = c(r.dn$p,     r.cn$p,     r.rn$p,     r.gn$p)
    )
  )
}


# Reads and rounds the edited rhandsontable
modified_table <- reactive({
  round(hot_to_r(req(input$dftbl1)))
})

# Extracts ca/co integer vectors from the table rows
ca_co <- reactive({
  tbl <- modified_table()
  list(
    ca = as.integer(tbl["Cases", ]),
    co = as.integer(tbl["Controls", ])
  )
})

# Live HWE computation (method driven by radioButtons input)
hwe_reactive <- reactive({
  d <- ca_co()
  hwe_pvalues(d$ca, d$co, method = input$hwe_method)
})
```


Data analysis
=====================================

Row {data-height=180}
-------------------------------------

### Data input

```{r}
global <- reactiveValues(df = df)

rHandsontableOutput("dftbl1")

output$dftbl1 <- renderRHandsontable({
  rhandsontable(global$df, selectCallback = TRUE, readOnly = FALSE, rowHeaderWidth = 80) %>%
    hot_context_menu(allowRowEdit = FALSE, allowColEdit = FALSE) %>%
    hot_validate_numeric(col = c("WT", "Het", "Hom"), min = 0, allowInvalid = FALSE)
})
```

### HWE test

```{r}
radioButtons("hwe_method",
             label = "Hardy-Weinberg test method",
             choices = c("Chi-square" = "chisq", "Exact (Wigginton 2005)" = "exact"),
             selected = "chisq")

```

### HWE results

```{r}
uiOutput("hwe_table")

output$hwe_table <- renderUI({
  d     <- ca_co()
  check <- check_data_validity(d$ca, d$co)
  if (!check$valid) return(p(check$msg))
  hwe          <- hwe_reactive()
  method_label <- if (input$hwe_method == "exact") "Exact H-W test" else "Chi-sq H-W test"
  stat_header  <- if (input$hwe_method == "exact") "\u2014" else "\u03c7\u00b2"
  rows <- list(
    c("Cases",
      if (is.na(hwe$t.ca)) "\u2014" else formatC(hwe$t.ca, format = "f", digits = 3),
      formatC(hwe$p.ca, format = "f", digits = 3)),
    c("Controls",
      if (is.na(hwe$t.co)) "\u2014" else formatC(hwe$t.co, format = "f", digits = 3),
      formatC(hwe$p.co, format = "f", digits = 3)),
    c("Both",
      formatC(hwe$t.2, format = "f", digits = 3),
      formatC(hwe$p.2, format = "f", digits = 3))
  )
  tagList(
    tags$b(method_label),
    tags$table(
      class = "table table-sm table-bordered",
      tags$thead(tags$tr(tags$th("Group"), tags$th(stat_header), tags$th("p"))),
      tags$tbody(lapply(rows, function(r) tags$tr(lapply(r, tags$td))))
    )
  )
})
```

### Save your work

```{r}
screenshotButton(selector = "body",
                 label    = "  Take a screenshot",
                 filename = "webassotest",
                 download = TRUE)

downloadButton("download_results", "  Download results (.csv)")

output$download_results <- downloadHandler(
  filename = function() paste0("webassotest_results_", Sys.Date(), ".csv"),
  content  = function(file) {
    d     <- ca_co()
    ca    <- d$ca
    co    <- d$co
    check <- check_data_validity(ca, co)
    req(check$valid)
    hwe <- hwe_pvalues(ca, co, method = input$hwe_method)
    res <- gtm(ca, co)
    hwe_df <- data.frame(
      Group    = c("Cases", "Controls", "Both"),
      HWE_stat = c(hwe$t.ca, hwe$t.co, hwe$t.2),
      HWE_p    = c(hwe$p.ca, hwe$p.co, hwe$p.2)
    )
    write.csv(rbind(
      cbind(Section = "Model results", res),
      cbind(Section = "HWE test", setNames(
        data.frame(hwe_df$Group, NA, NA, NA, hwe_df$HWE_stat, hwe_df$HWE_p),
        names(res)))
    ), file, row.names = FALSE)
  }
)
```


Row
-------------------------------------

### Result

```{r}
renderPlot({
  d     <- ca_co()
  ca    <- d$ca
  co    <- d$co
  check <- check_data_validity(ca, co)
  validate(need(check$valid, check$msg))
  gtm(ca, co)
})
```


About {#about}
==========================

Column {data-width=600}
-----------------------------------------------------------------------

### About Web-Assotest

Web-Assotest is a browser-based tool for analysing the association between a genetic variant and a binary outcome (cases vs. controls) using a 3×2 genotype count table. It fits four standard genetic models — **dominant**, **recessive**, **co-dominant**, and **genotype** — and displays odds ratios with 95% confidence intervals and likelihood-ratio test statistics for each, allowing the user to assess which models are consistent with the data.

The tool also tests for departure from **Hardy-Weinberg equilibrium (HWE)** in cases, controls, and both combined, using either a chi-square test or the exact test of Wigginton, Cutler & Abecasis (2005).

For an overview and explanation of the models, see the [course slides](https://ekstroem.com/download/assotestslides.pdf) on case-control studies and genetic association analysis. Web-Assotest is [widely used in scientific publications](http://scholar.google.com/scholar?q=web-assotest).

Web-Assotest uses [R](http://www.r-project.org/) for all computations. See the [R Primer book](http://www.rprimer.dk/) for an introduction to statistical analysis in R.

**Note:** cells with zero counts are handled using Fisher's exact test for the affected contrasts. Take care when many cells contain zeros.

### References

Wigginton JE, Cutler DJ, Abecasis GR (2005). A Note on Exact Tests of Hardy-Weinberg Equilibrium. *American Journal of Human Genetics*, 76(5):887–893.
