---
title: "WEB - Assotest"
date: 'Page loaded: `r format(as.POSIXct(format(Sys.time())), "%b %d, %Y")`'
output:
  flexdashboard::flex_dashboard:
    navbar:
      - { title: "v0.99", align: right }
    orientation: rows
    vertical_layout: scroll
    social: menu
    theme:
      version: 4
      base_font:
        google: Prompt
      heading_font:
        google: Sen
runtime: shiny
---

<style type="text/css">
  body {
  font-size: 16pt;
}


h1,h2,h3,h4,h5,h6 {
  font-size: 24pt;
}

.navbar-brand {
  color:white;
  font-size: 32px;
}

div.navbar-header {
  font-size: 28pt;
}

div.chart-title {
  font-size: 28pt;
}

@media (max-width: 1200px) {
  body           { font-size: 13pt; }
  h1,h2,h3,h4,h5,h6 { font-size: 18pt; }
  .navbar-brand  { font-size: 24px; }
  div.navbar-header, div.chart-title { font-size: 20pt; }
}

@media (max-width: 768px) {
  body           { font-size: 11pt; }
  h1,h2,h3,h4,h5,h6 { font-size: 15pt; }
  .navbar-brand  { font-size: 18px; }
  div.navbar-header, div.chart-title { font-size: 16pt; }
}

/* Center the page-loaded date in the navbar */
.navbar-author {
  position: absolute;
  left: 50%;
  transform: translateX(-50%);
  color: rgba(255,255,255,0.8);
  line-height: 50px;
  white-space: nowrap;
  pointer-events: none;
}

/* Hide the About page tab from the left-side tab navigation only */

/* Left-align rhandsontable row header labels */
.handsontable tbody th {
  text-align: left !important;
}

/* Shrink rhandsontable output container to its table content height */
#dftbl1 {
  height: auto !important;
}

.input-panel,
.analysis-panel,
.hwe-panel {
  flex: 2 1 0% !important;
  max-width: none !important;
  min-width: 0 !important;
}

/* Save buttons: centred column, equal width */
.save-buttons {
  display: flex;
  flex-direction: column;
  align-items: center;
  justify-content: center;
  height: 100%;
  gap: 10px;
}
.save-buttons .btn {
  width: 100% !important;
  flex: 1;
  display: flex !important;
  align-items: center;
  justify-content: center;
  font-size: 16pt !important;
  background-color: #c8e6c9 !important;
  border-color: #a5d6a7 !important;
  color: #1b5e20 !important;
}

.save-panel {
  flex: 1 1 0% !important;
  max-width: none !important;
  min-width: 0 !important;
}
</style>

<script>
$(document).ready(function() {
  // Remove the auto-generated About tab (no glyphicon); keep the YAML navbar item (has glyphicon)
  $('.navbar-nav a[href="#about"]').each(function() {
    if ($(this).find('.glyphicon').length === 0) {
      $(this).closest('li').remove();
    }
  });
});
</script>


```{r setup, include=FALSE}
# --------------------------------------------------------------------
# App architecture overview
#
# This is a single-file Shiny flexdashboard. UI and server logic are
# both defined in R chunks; there is no separate ui.R / server.R.
#
# Layout (orientation: rows, vertical_layout: scroll):
#   Row 1 — three panels side by side (1/6 | 2/3 | 1/6 of width):
#     · Data input    — rhandsontable for genotype counts
#     · HWE           — test method selector + results table
#     · Save work     — screenshot and CSV download buttons
#   Row 2 — full-width model-comparison diagram
#
# Reactive chain:
#   input$dftbl1 (table edits)
#     └─► modified_table  — rounds raw rhandsontable values
#           └─► ca_co     — extracts ca/co integer vectors
#                 ├─► hwe_reactive ──► output$hwe_table  (HWE panel)
#                 └─────────────────► output$result_plot (diagram)
#
# Pure functions defined below (before reactives so they are available
# to all outputs): SNPHWE, check_data_validity, hwe_pvalues, maf,
# ci.mat, fmt_p, p_lab, fmt_ci, fisher_or, allelic_tests, gtm.
# --------------------------------------------------------------------

library("flexdashboard")
library("rhandsontable")
library("shinyscreenshot")
library("logistf")

# Application version, written to the CSV export.
# Keep in sync with the "v0.99" navbar item in the YAML header.
app_version <- "0.99"

# Default genotype count table — rows: Cases/Controls, cols: WT/Het/Hom
df <- data.frame(
  WT  = c(10L, 10L),
  Het = c(20L, 20L),
  Hom = c(30L, 30L)
)
rownames(df) <- c("Cases", "Controls")

# --------------------------------------------------------------------
# SNPHWE(obs_hets, obs_hom1, obs_hom2)
# Exact test of Hardy-Weinberg Equilibrium.
# Reference: Wigginton JE, Cutler DJ, Abecasis GR (2005). A Note on
# Exact Tests of Hardy-Weinberg Equilibrium. Am J Hum Genet 76:887-893.
#
# obs_hom1, obs_hom2: homozygote counts (order does not matter;
#   the algorithm identifies the rare homozygote internally via min/max).
# obs_hets: heterozygote count.
# Returns the two-sided p-value, or -1.0 on invalid (negative) input.
#
# Algorithm: enumerates the conditional distribution of heterozygote
# counts given fixed marginals, computing probabilities by recurrence
# outward from the mode to avoid overflow.
# --------------------------------------------------------------------
SNPHWE <- function(obs_hets, obs_hom1, obs_hom2) {
  if (obs_hom1 < 0 || obs_hom2 < 0 || obs_hets < 0)
    return(-1.0)

  N       <- obs_hom1 + obs_hom2 + obs_hets
  obs_homr <- min(obs_hom1, obs_hom2)
  obs_homc <- max(obs_hom1, obs_hom2)
  rare    <- obs_homr * 2 + obs_hets

  probs <- rep(0, 1 + rare)

  mid <- floor(rare * (2 * N - rare) / (2 * N))
  if ((mid %% 2) != (rare %% 2)) mid <- mid + 1

  probs[mid + 1] <- 1.0
  mysum <- 1.0

  # Probabilities from midpoint downward
  curr_hets <- mid
  curr_homr <- (rare - mid) / 2
  curr_homc <- N - curr_hets - curr_homr

  while (curr_hets >= 2) {
    probs[curr_hets - 1] <- probs[curr_hets + 1] * curr_hets *
      (curr_hets - 1.0) / (4.0 * (curr_homr + 1.0) * (curr_homc + 1.0))
    mysum     <- mysum + probs[curr_hets - 1]
    curr_hets <- curr_hets - 2
    curr_homr <- curr_homr + 1
    curr_homc <- curr_homc + 1
  }

  # Probabilities from midpoint upward
  curr_hets <- mid
  curr_homr <- (rare - mid) / 2
  curr_homc <- N - curr_hets - curr_homr

  while (curr_hets <= rare - 2) {
    probs[curr_hets + 3] <- probs[curr_hets + 1] * 4.0 * curr_homr *
      curr_homc / ((curr_hets + 2.0) * (curr_hets + 1.0))
    mysum     <- mysum + probs[curr_hets + 3]
    curr_hets <- curr_hets + 2
    curr_homr <- curr_homr - 1
    curr_homc <- curr_homc - 1
  }

  target <- probs[obs_hets + 1]
  # Relative tolerance so floating-point ties with the observed table count
  p <- min(1.0, sum(probs[probs <= target * (1 + 1e-7)]) / mysum)
  return(p)
}


# --------------------------------------------------------------------
# check_data_validity(ca, co)
# ca, co: integer vectors of length 3 (WT, Het, Hom counts).
# Returns list(valid = TRUE/FALSE, msg = character).
# Rejects tables where all cases or all controls are zero, or where
# fewer than 2 genotype groups have any observations (models cannot
# be identified with only one non-empty column).
# --------------------------------------------------------------------
check_data_validity <- function(ca, co) {
  if (sum(ca + co == 0) > 1) return(list(valid = FALSE,
    msg = "At least two genotype groups must have observations."))
  if (sum(ca) == 0) return(list(valid = FALSE,
    msg = "There are no cases. Please enter case counts."))
  if (sum(co) == 0) return(list(valid = FALSE,
    msg = "There are no controls. Please enter control counts."))
  return(list(valid = TRUE, msg = ""))
}


# --------------------------------------------------------------------
# hwe_pvalues(ca, co, method)
# ca, co: integer vectors of length 3. method: "chisq" or "exact".
# Returns a list with test statistics (t.ca, t.co, t.2) and p-values
# (p.ca, p.co, p.2) for cases, controls, and both combined.
#
# For method = "exact": individual statistics are NA (SNPHWE returns
# only a p-value); the combined p-value uses Fisher's method —
# -2*(log(p.ca) + log(p.co)) ~ chi-sq(4).
# For method = "chisq": combined statistic is the sum of the two
# chi-square values ~ chi-sq(2).
# --------------------------------------------------------------------
hwe_pvalues <- function(ca, co, method = "chisq") {

  # Chi-square HWE test for a single group x = c(WT, Het, Hom)
  hwe_chisq <- function(x) {
    f   <- (2 * x[1] + x[2]) / (2 * sum(x))   # reference allele frequency
    # Monomorphic group: expected counts are zero, HWE holds trivially
    if (f %in% c(0, 1)) return(list(stat = 0, p = 1))
    e   <- c(f^2, 2 * f * (1 - f), (1 - f)^2) * sum(x)
    tst <- sum((x - e)^2 / e)
    p   <- pchisq(tst, 1, lower.tail = FALSE)
    list(stat = tst, p = p)
  }

  if (method == "exact") {
    p.ca  <- SNPHWE(ca[2], ca[1], ca[3])
    p.co  <- SNPHWE(co[2], co[1], co[3])
    # Combined: sum of -2*log(p) ~ chi-sq(4) (Fisher's method)
    t.2   <- -2 * (log(p.ca) + log(p.co))
    p.2   <- pchisq(t.2, 4, lower.tail = FALSE)
    t.ca  <- NA
    t.co  <- NA
  } else {
    r.ca  <- hwe_chisq(ca)
    r.co  <- hwe_chisq(co)
    t.ca  <- r.ca$stat;  p.ca <- r.ca$p
    t.co  <- r.co$stat;  p.co <- r.co$p
    t.2   <- t.ca + t.co
    p.2   <- pchisq(t.2, 2, lower.tail = FALSE)
  }

  list(t.ca = t.ca, p.ca = p.ca,
       t.co = t.co, p.co = p.co,
       t.2  = t.2,  p.2  = p.2)
}


# --------------------------------------------------------------------
# maf(x)
# x: integer vector of length 3 (WT, Het, Hom counts).
# Returns the minor allele frequency, or NA when total count is zero.
# --------------------------------------------------------------------
maf <- function(x) {
  n <- sum(x)
  if (n == 0) return(NA)
  p <- (2 * x[1] + x[2]) / (2 * n)   # reference allele frequency
  min(p, 1 - p)
}


# --------------------------------------------------------------------
# ci.mat(alpha)
# Returns a 2x3 matrix for simultaneous Wald CI computation.
# Post-multiply a [estimate, se] matrix by ci.mat() to obtain
# [estimate, lower, upper] in one step — e.g.:
#   exp(summary(fit)$coef[, 1:2] %*% ci.mat())
# gives the OR with 95% CIs directly from log-OR and its SE.
# --------------------------------------------------------------------
ci.mat <- function(alpha = 0.05) {
  ciM <- rbind(c(1, 1, 1), qnorm(1 - alpha / 2) * c(0, -1, 1))
  colnames(ciM) <- c("Estimate",
    paste0(formatC(100 *    alpha / 2,  format = "f", digits = 1), "%"),
    paste0(formatC(100 * (1 - alpha / 2), format = "f", digits = 1), "%"))
  ciM
}


# --------------------------------------------------------------------
# Formatting helpers.
# fmt_p(p):   p-value as text; small values in scientific notation
#             ("1.2e-10") instead of "0.000", floored at "<1e-300".
# p_lab(p):   "p=0.123" or "p<1e-300" for plot labels.
# fmt_ci(x):  x = c(estimate, lower, upper) -> "1.23 ( 0.90 ; 1.70 )".
# --------------------------------------------------------------------
fmt_p <- function(p) format.pval(p, digits = 3, eps = 1e-300)

p_lab <- function(p) {
  s <- fmt_p(p)
  if (startsWith(s, "<")) paste0("p", s) else paste0("p=", s)
}

fmt_ci <- function(x) {
  f <- formatC(x, format = "f", digits = 2)
  paste(f[1], "(", f[2], ";", f[3], ")")
}


# --------------------------------------------------------------------
# fisher_or(tab)
# tab: 2x2 matrix, rows Controls/Cases, columns reference/exposed.
# Returns c(estimate, lower, upper) from Fisher's exact test — the
# conditional MLE odds ratio, used when a cell is zero.
# --------------------------------------------------------------------
fisher_or <- function(tab) {
  ft <- fisher.test(tab)
  c(ft[["estimate"]], ft[["conf.int"]])
}


# --------------------------------------------------------------------
# allelic_tests(ca, co)
# ca, co: integer vectors of length 3 (WT, Het, Hom counts).
# Returns a data frame with two 1-df tests:
#   Cochran-Armitage trend test (prop.trend.test, scores 1/2/3) —
#     valid without HWE.
#   Allelic test — 2x2 chi-square on allele counts, with the allelic
#     OR (Het/Hom allele vs WT allele) and Wald 95% CI; Fisher's exact
#     OR/CI when an allele cell is zero. Assumes HWE.
# --------------------------------------------------------------------
allelic_tests <- function(ca, co) {
  trend  <- suppressWarnings(prop.trend.test(ca, ca + co))
  alle   <- function(x) c(2 * x[1] + x[2], x[2] + 2 * x[3])
  tab    <- rbind(alle(co), alle(ca))   # rows Controls/Cases, cols ref/alt allele
  # Allele columns can be empty (monomorphic sample) — no test then
  chi    <- if (all(colSums(tab) > 0))
    suppressWarnings(chisq.test(tab, correct = FALSE)) else list(statistic = NA, p.value = NA)
  or <- if (any(tab == 0)) fisher_or(tab) else {
    lor <- log(tab[1, 1] * tab[2, 2] / (tab[1, 2] * tab[2, 1]))
    se  <- sqrt(sum(1 / tab))
    exp(lor + c(0, -1, 1) * qnorm(0.975) * se)
  }
  data.frame(
    Test     = c("Cochran-Armitage", "Allelic"),
    df       = c(1, 1),
    Chisq    = c(unname(trend$statistic), unname(chi$statistic)),
    p_value  = c(trend$p.value, chi$p.value),
    OR       = c(NA, or[1]),
    CI_lower = c(NA, or[2]),
    CI_upper = c(NA, or[3])
  )
}


# --------------------------------------------------------------------
# gtm(ca, co, flev, fnam, main, firth)
# ca, co: integer vectors of length 3 (WT, Het, Hom counts).
# flev: genotype level labels. fnam/main: plot title.
# firth: if TRUE, use Firth's penalized logistic regression (logistf
#   package); CIs are profile-likelihood-based and no Fisher fallback
#   is applied (Firth handles separation natively).
#
# Fits five nested logistic regression models:
#   nul: intercept only (no association)
#   gen: factor(WT/Het/Hom) — genotype model, 2 df
#   dom: factor(WT vs Het+Hom) — dominant model, 1 df
#   cod: linear score 0/1/2 — additive (log-additive) model, 1 df
#   rec: factor(WT+Het vs Hom) — recessive model, 1 df
#
# Draws a model-comparison diagram using base graphics. Lines connect
# nested models; lwd=8 when the simpler model cannot be rejected at
# p >= 0.05 (compatible), lwd=2 otherwise (rejected).
#
# When any cell in a contrast is zero, Fisher's exact test replaces
# the logistic-regression OR and CI for that contrast.
#
# Returns invisibly list(models, tests): per-contrast ORs/CIs with AIC,
# and all likelihood-ratio tests (used by the AIC table and CSV download).
# --------------------------------------------------------------------

gtm <- function(ca, co,
                flev = c("WT", "Het", "Hom"),
                fnam = " ",
                main = fnam,
                firth = FALSE) {

  rlev <- c(rep(paste(flev[1:2], collapse = "/"), 2), flev[3])
  dlev <- c(flev[1], rep(paste(flev[2:3], collapse = "/"), 2))
  f.g  <- factor(flev, levels = flev)
  f.d  <- factor(dlev, levels = unique(dlev))
  v.c  <- 0:2
  f.r  <- factor(rlev, levels = unique(rlev))

  old <- par(mar = c(0, 0, 2, 0), cex = 1.6)
  on.exit(par(old))
  plot(NA, type = "n", axes = FALSE, xlab = "", ylab = "",
       xlim = c(-10, 100), ylim = c(-5, 105), main = main)

  # Guard: warn if data are marginal but still plottable
  if (sum(c(ca, co) == 0) > 3)
    text(-10, 80, "Warning: many zero cells — interpret with caution.",
         font = 2, adj = 0)

  if (firth) {
    # Firth's penalized logistic regression (logistf).
    # logistf requires individual binary observations, not grouped counts.
    y   <- c(rep(1L, sum(ca)), rep(0L, sum(co)))
    grp <- c(rep(1:3, ca), rep(1:3, co))
    # Predictor vectors shared across models so anova() term-nesting check passes.
    # x1 = dominant indicator (Het|Hom vs WT)
    # x2 = recessive indicator (Hom vs WT|Het)
    # xa = additive score (0/1/2 allele copies)
    x1 <- as.integer(grp > 1)
    x2 <- as.integer(grp == 3)
    xa <- x1 + x2
    nul    <- logistf(y ~ 1)
    gen    <- logistf(y ~ factor(flev[grp], levels = flev))  # for OR extraction
    gen_dr <- logistf(y ~ x1 + x2)   # dom/rec/nul anova parent; same loglik as gen
    gen_ca <- logistf(y ~ xa + x2)   # cod anova parent; same loglik as gen
    dom    <- logistf(y ~ x1)        # nested in gen_dr
    cod    <- logistf(y ~ xa)        # nested in gen_ca
    rec    <- logistf(y ~ x2)        # nested in gen_dr
    
    # Extract the test statistic and p-value. Statistics within rounding
    # error of zero are set to 0 so formatC does not print "-0.000".
    plr  <- function(a) {
      if (a$chisq<0.0000001)
        a$chisq <- 0
      list(chisq = a$chisq, p = a$pval)
    }
    r.dn <- plr(anova(nul, dom))
    r.cn <- plr(anova(nul, cod))
    r.rn <- plr(anova(nul, rec))
    r.gd <- plr(anova(dom, gen_dr))
    r.gc <- plr(anova(cod, gen_ca))
    r.gr <- plr(anova(rec, gen_dr))
    r.gn <- plr(anova(nul, gen_dr))
    or_firth <- function(fit) {
      cf <- fit$coefficients[-1]
      matrix(c(exp(cf), exp(fit$ci.lower[-1]), exp(fit$ci.upper[-1])),
             ncol = 3, dimnames = list(NULL, c("Estimate", "2.5%", "97.5%")))
    }
    OR.g <- or_firth(gen)
    OR.d <- or_firth(dom)
    OR.c <- or_firth(cod)
    OR.r <- or_firth(rec)
    aic_vals <- sapply(list(nul, dom, cod, rec, gen), function(fit)
      2 * length(fit$coefficients) - 2 * fit$loglik[2])
  } else {
    # Standard grouped-binomial logistic regression via glm
    nul <- glm(cbind(ca, co) ~ 1,   family = binomial)
    gen <- glm(cbind(ca, co) ~ f.g, family = binomial)
    dom <- glm(cbind(ca, co) ~ f.d, family = binomial)
    cod <- glm(cbind(ca, co) ~ v.c, family = binomial)
    rec <- glm(cbind(ca, co) ~ f.r, family = binomial)
    t.dn  <- anova(nul, dom, test = "Chisq")
    t.cn  <- anova(nul, cod, test = "Chisq")
    t.rn  <- anova(nul, rec, test = "Chisq")
    t.gd  <- anova(dom, gen, test = "Chisq")
    t.gc  <- anova(cod, gen, test = "Chisq")
    t.gr  <- anova(rec, gen, test = "Chisq")
    t.gn  <- anova(nul, gen, test = "Chisq")
    extract <- function(a) list(chisq = abs(a[2, 4]), p = a[2, 5])
    r.dn <- extract(t.dn); r.cn <- extract(t.cn); r.rn <- extract(t.rn)
    r.gd <- extract(t.gd); r.gc <- extract(t.gc); r.gr <- extract(t.gr)
    r.gn <- extract(t.gn)
    OR.g <- exp(summary(gen)$coef[-1, 1:2, drop = FALSE] %*% ci.mat())
    OR.d <- exp(summary(dom)$coef[-1, 1:2, drop = FALSE] %*% ci.mat())
    OR.c <- exp(summary(cod)$coef[-1, 1:2, drop = FALSE] %*% ci.mat())
    OR.r <- exp(summary(rec)$coef[-1, 1:2, drop = FALSE] %*% ci.mat())
    aic_vals <- c(AIC(nul), AIC(dom), AIC(cod), AIC(rec), AIC(gen))
  }
  
  # Pad OR matrices in case of singularities (complete separation)
  if (nrow(OR.g) < 2) OR.g <- rbind(OR.g, rep(NA, 3))
  if (nrow(OR.d) == 0) OR.d <- matrix(NA, 1, 3)
  if (nrow(OR.r) == 0) OR.r <- matrix(NA, 1, 3)

  # Fall back to Fisher's exact test when zeros are present.
  # Skipped for Firth regression — it handles separation natively.
  if (!firth) {
    gt1 <- rbind(co, ca)[, -3]   # WT vs Het
    gt2 <- rbind(co, ca)[, -2]   # WT vs Hom
    if (any(gt1 == 0)) OR.g[1, ] <- fisher_or(gt1)
    if (any(gt2 == 0)) OR.g[2, ] <- fisher_or(gt2)
    dot <- rbind(co, ca) %*% cbind(c(1, 0, 0), c(0, 1, 1))  # WT vs Het+Hom
    if (any(dot == 0)) OR.d[1, ] <- fisher_or(dot)
    if (sum(wh <- (ca + co > 0)) < 3) {
      cot <- rbind(co, ca)[, wh]
      if (any(cot == 0)) {
        OR.c[1, ] <- fisher_or(cot)
        # When the Het column is missing, the additive OR is per 2 copies;
        # take the square root to recover the per-allele OR.
        if (!wh[2]) OR.c <- sqrt(OR.c)
      }
    }
    rct <- rbind(co, ca) %*% cbind(c(1, 1, 0), c(0, 0, 1))  # WT+Het vs Hom
    if (any(rct == 0)) OR.r[1, ] <- fisher_or(rct)
  }

  # Genotype model — top centre
  text(50, 98, "Genotype model", font = 2)
  text(50, 95, paste("OR(", flev[2], "vs.", flev[1], "):", fmt_ci(OR.g[1, ])))
  text(50, 92, paste("OR(", flev[3], "vs.", flev[1], "):", fmt_ci(OR.g[2, ])))

  # Dominant model — left
  text(0, 53, "Dominant", adj = c(0, 0.5), font = 2)
  text(0, 50, paste("OR(", dlev[2], "vs.", dlev[1], "):"), adj = c(0, 0.5))
  text(0, 47, fmt_ci(OR.d[1, ]), adj = c(0, 0.5))

  # Additive (log-additive) model — centre
  text(50, 53, "Additive", adj = c(0.5, 0.5), font = 2)
  text(50, 50, paste("OR(", flev[3], "vs.", flev[2], ")=",
                     "OR(", flev[2], "vs.", flev[1], "):"), adj = c(0.5, 0.5))
  text(50, 47, fmt_ci(OR.c[1, ]), adj = c(0.5, 0.5))

  # Recessive model — right
  text(100, 53, "Recessive", adj = c(1, 0.5), font = 2)
  text(100, 50, paste("OR(", rlev[3], "vs.", rlev[1], "):"), adj = c(1, 0.5))
  text(100, 47, fmt_ci(OR.r[1, ]), adj = c(1, 0.5))

  # Null model — bottom centre
  text(50, 8, "Null model", adj = c(0.5, 0.5), font = 2)
  text(50, 5, "OR = 1", adj = c(0.5, 0.5))
  text(50, 2, substitute(chi^2 * rest,
    list(rest = paste0("(2)=", formatC(r.gn$chisq, format = "f", digits = 3),
                       ", ", p_lab(r.gn$p)))),
    adj = c(0.5, 0.5))

  # Connecting lines: lwd=8 when p >= 0.05 (simpler model not rejected),
  # lwd=2 when rejected. Thick lines indicate compatible simplifications.
  lwd_for <- function(p) if (is.na(p) || p >= 0.05) 8 else 2

  lines(c(40, 10), c(90, 55), lwd = lwd_for(r.gd$p))  # Genotype  -> Dominant
  lines(c(10, 40), c(45, 10), lwd = lwd_for(r.dn$p))  # Dominant  -> Null
  lines(c(50, 50), c(10, 45), lwd = lwd_for(r.cn$p))  # Null      -> Additive
  lines(c(50, 50), c(55, 90), lwd = lwd_for(r.gc$p))  # Additive  -> Genotype
  lines(c(60, 90), c(90, 55), lwd = lwd_for(r.gr$p))  # Genotype  -> Recessive
  lines(c(90, 60), c(45, 10), lwd = lwd_for(r.rn$p))  # Recessive -> Null

  legend("bottomleft", lwd = c(8, 2), bty = "n",
         legend = c("Not rejected (p ≥ 0.05)", "Rejected (p < 0.05)"))

  rect(0, 25, 100, 30, col = "white", border = "white")
  rect(0, 70, 100, 75, col = "white", border = "white")

  # Test statistics along connecting lines
  # Order matches line segments: gd, dn, cn, gc, gr, rn
  results_list <- list(r.gd, r.dn, r.cn, r.gc, r.gr, r.rn)

  results_list <- lapply(results_list, function(r) {
    if (is.na(r$p) && !is.na(r$chisq) && r$chisq < 0.000001) r$p <- 1
    r
  })

  # Set the positions of the test statistics on the lines
  xt <- c(30, 30, 50, 50, 70, 70)
  yt <- c(72.5, 27.5, 27.5, 72.5, 72.5, 27.5)
  at <- c(0.9, 0.9, 0.5, 0.5, 0.1, 0.1)

  for (i in 1:6)
    text(xt[i], yt[i],
         substitute(chi^2 == rest,
           list(rest = paste0(formatC(results_list[[i]]$chisq, format = "f", digits = 3),
                              ", ", p_lab(results_list[[i]]$p)))),
         adj = c(at[i], 0.5))

  # Return tidy summaries for the AIC table and the CSV download.
  # models: one row per OR contrast (Genotype has two); AIC per model.
  # tests:  every likelihood-ratio test shown on the diagram.
  or_rows <- rbind(rep(NA, 3), OR.d[1, ], OR.c[1, ], OR.r[1, ], OR.g[1, ], OR.g[2, ])
  invisible(list(
    models = data.frame(
      Model    = c("Null", "Dominant", "Additive", "Recessive", "Genotype", "Genotype"),
      Contrast = c(NA, paste(dlev[2], "vs", dlev[1]), "per allele copy",
                   paste(rlev[3], "vs", rlev[1]),
                   paste(flev[2], "vs", flev[1]), paste(flev[3], "vs", flev[1])),
      OR       = or_rows[, 1],
      CI_lower = or_rows[, 2],
      CI_upper = or_rows[, 3],
      AIC      = aic_vals[c(1:5, 5)]
    ),
    tests = data.frame(
      Comparison = c("Dominant vs Null", "Additive vs Null", "Recessive vs Null",
                     "Genotype vs Null", "Genotype vs Dominant",
                     "Genotype vs Additive", "Genotype vs Recessive"),
      df      = c(1, 1, 1, 2, 1, 1, 1),
      Chisq   = c(r.dn$chisq, r.cn$chisq, r.rn$chisq, r.gn$chisq,
                  r.gd$chisq, r.gc$chisq, r.gr$chisq),
      p_value = c(r.dn$p, r.cn$p, r.rn$p, r.gn$p, r.gd$p, r.gc$p, r.gr$p)
    )
  ))
}


# Reads the edited rhandsontable and coerces to non-negative integers.
# Server-side backstop for hot_validate_numeric(min=0) — belt and braces
# in case a value reaches here without going through client-side validation.
modified_table <- reactive({
  tbl <- hot_to_r(req(input$dftbl1))
  tbl[] <- lapply(tbl, function(x) as.integer(pmax(0, round(x))))
  tbl
})

# Extracts ca/co integer vectors from the table rows
ca_co <- reactive({
  tbl <- modified_table()
  list(
    ca = as.integer(tbl["Cases", ]),
    co = as.integer(tbl["Controls", ])
  )
})

# Live HWE computation (method driven by radioButtons input)
hwe_reactive <- reactive({
  d <- ca_co()
  hwe_pvalues(d$ca, d$co, method = input$hwe_method)
})
```



Data analysis
=====================================

Row {data-height=320}
-------------------------------------

### Data input {data-width=1 .input-panel}

```{r}
# global holds the table state used to seed the rhandsontable widget.
# Edits are captured via input$dftbl1 (hot_to_r); global is only replaced
# by the Flip/Reset buttons. nonce forces a re-render even when the new
# df is identical to the old one (e.g. Reset after editing the table).
global <- reactiveValues(df = df, nonce = 0)
set_table <- function(new_df) {
  global$df    <- new_df
  global$nonce <- isolate(global$nonce) + 1
}

rHandsontableOutput("dftbl1")
uiOutput("row_pct")

tags$div(style = "display:flex; gap:6px; margin:6px 0;",
  actionButton("flip", "Flip allele coding", class = "btn-sm",
               title = "Swap WT and Hom columns (and their labels)"),
  actionButton("reset", "Reset", class = "btn-sm",
               title = "Restore the default table and labels"))

# Row percentages per genotype, so case/control distributions are easy
# to compare at a glance.
output$row_pct <- renderUI({
  d   <- ca_co()
  pct <- function(x) if (sum(x) == 0) "\u2014" else
    paste0(round(100 * x / sum(x)), "%", collapse = " / ")
  tags$small(class = "text-muted",
    tags$div("Cases: ",    pct(d$ca)),
    tags$div("Controls: ", pct(d$co)))
})

# Flip: reverse genotype order so the other allele is the reference.
# Dominant/recessive swap meaning, so labels are swapped too.
observeEvent(input$flip, {
  tbl <- modified_table()[, 3:1]
  names(tbl) <- c("WT", "Het", "Hom")
  set_table(tbl)
  updateTextInput(session, "lbl_wt",  value = input$lbl_hom)
  updateTextInput(session, "lbl_hom", value = input$lbl_wt)
})

observeEvent(input$reset, {
  set_table(df)
  updateTextInput(session, "lbl_wt",  value = "WT")
  updateTextInput(session, "lbl_het", value = "Het")
  updateTextInput(session, "lbl_hom", value = "Hom")
})

# Optional genotype/allele labels shown on the model-comparison diagram
# instead of the generic WT/Het/Hom. Blank fields fall back to the default.
# Plain flex div (not fluidRow/column) — this panel is a flex item with
# max-width:none/min-width:0, and fluidRow's negative-margin grid overflows
# that kind of parent instead of a bootstrap .container, which corrupts the
# panel's layout and, via the resize observer, the plot below it.
# Built with raw tags rather than textInput() so the label sits beside
# (not above) its box — textInput always stacks label above input.
tags$div(
  style = "display:flex; flex-direction:column; gap:8px;",
  tags$div(style = "display:flex; align-items:center; gap:6px;",
    tags$label("WT label",  `for` = "lbl_wt",  style = "margin:0; white-space:nowrap;"),
    tags$input(id = "lbl_wt",  type = "text", class = "form-control", value = "WT")),
  tags$div(style = "display:flex; align-items:center; gap:6px;",
    tags$label("Het label", `for` = "lbl_het", style = "margin:0; white-space:nowrap;"),
    tags$input(id = "lbl_het", type = "text", class = "form-control", value = "Het")),
  tags$div(style = "display:flex; align-items:center; gap:6px;",
    tags$label("Hom label", `for` = "lbl_hom", style = "margin:0; white-space:nowrap;"),
    tags$input(id = "lbl_hom", type = "text", class = "form-control", value = "Hom"))
)

uiOutput("label_warning")

# Genotype labels used by gtm() for plot text; blank entries fall back to
# the WT/Het/Hom default so the diagram never shows an empty label.
flev_labels <- reactive({
  lbl <- c(input$lbl_wt, input$lbl_het, input$lbl_hom)
  dflt <- c("WT", "Het", "Hom")
  ifelse(is.na(lbl) | trimws(lbl) == "", dflt, trimws(lbl))
})

# Warn (non-blocking) when two or three resolved labels collide — the
# diagram would otherwise show an ambiguous contrast like "OR(Mut vs Mut)".
output$label_warning <- renderUI({
  lbl <- flev_labels()
  if (anyDuplicated(tolower(lbl)) == 0) return(NULL)
  tags$p(class = "text-danger", style = "margin-top:4px; margin-bottom:0;",
         "Genotype labels must be unique — duplicate labels will be confusing on the diagram.")
})

# Render the editable 2x3 genotype count table.
# Row/column editing is disabled — only cell values can be changed.
# Numeric validation enforces non-negative values and prevents non-numeric input.
# rowHeaderWidth = 80 ensures "Cases" / "Controls" labels are not clipped.
output$dftbl1 <- renderRHandsontable({
  global$nonce
  rhandsontable(global$df, readOnly = FALSE, rowHeaderWidth = 80) %>%
    hot_context_menu(allowRowEdit = FALSE, allowColEdit = FALSE) %>%
    hot_validate_numeric(col = c("WT", "Het", "Hom"), min = 0, allowInvalid = FALSE) %>%
    hot_cols(colWidths = 45)
})
```

### Analysis options {data-width=1 .analysis-panel}

```{r}
# Regression method (Firth vs standard) and HWE test method, grouped
# together since both control how results are computed.
checkboxInput("use_firth", "Use Firth's penalized regression", value = FALSE)

radioButtons("hwe_method",
             label = "HWE test method",
             choices = c("Chi-square" = "chisq", "Exact (Wigginton 2005)" = "exact"),
             selected = "chisq")
```

### Hardy-Weinberg equilibrium {data-width=1 .hwe-panel}

```{r}
# hwe_method radio button lives in the Analysis options panel; its choice
# propagates to hwe_reactive, which drives both the table display here and
# the HWE section of the CSV download.

# uiOutput placeholder; filled by output$hwe_table below.
uiOutput("hwe_table")

# Renders the HWE results as an HTML table with four columns:
#   Group | MAF | test statistic | p-value
# Rows: Cases, Controls, Both (combined via Fisher's method or sum of chi-sq).
# Returns an error message paragraph instead of the table when input is invalid.
# The stat column header and "Both" row statistic differ by method:
#   chi-sq:  individual chi-sq stats; combined = sum ~ chi-sq(2)
#   exact:   individual stats are NA (shown as \u2014); combined uses Fisher's
#            method: -2*(log(p_ca) + log(p_co)) ~ chi-sq(4)
output$hwe_table <- renderUI({
  d     <- ca_co()
  check <- check_data_validity(d$ca, d$co)
  if (!check$valid) return(p(check$msg))
  hwe          <- hwe_reactive()
  method_label <- if (input$hwe_method == "exact") "Exact H-W test" else "Chi-sq H-W test"
  stat_header  <- if (input$hwe_method == "exact") "\u2014" else "\u03c7\u00b2"
  fmt_maf <- function(x) if (is.na(x)) "\u2014" else formatC(x, format = "f", digits = 3)
  rows <- list(
    c("Cases",
      fmt_maf(maf(d$ca)),
      if (is.na(hwe$t.ca)) "\u2014" else formatC(hwe$t.ca, format = "f", digits = 3),
      fmt_p(hwe$p.ca)),
    c("Controls",
      fmt_maf(maf(d$co)),
      if (is.na(hwe$t.co)) "\u2014" else formatC(hwe$t.co, format = "f", digits = 3),
      fmt_p(hwe$p.co)),
    c("Both",
      fmt_maf(maf(d$ca + d$co)),
      formatC(hwe$t.2, format = "f", digits = 3),
      fmt_p(hwe$p.2))
  )
  tagList(
    tags$b(method_label),
    tags$table(
      class = "table table-sm table-bordered",
      tags$thead(tags$tr(tags$th("Group"), tags$th("MAF"), tags$th(stat_header), tags$th("p"))),
      tags$tbody(lapply(rows, function(r) tags$tr(lapply(r, tags$td))))
    )
  )
})
```

### Save your work {data-width=1 .save-panel}

```{r}
# Two full-width buttons stacked vertically and centred via .save-buttons CSS.
# screenshotButton captures the entire page body as a PNG.
# downloadButton triggers output$download_results below.
tags$div(
  class = "save-buttons",
  screenshotButton(selector = "body",
                   label    = " Take a screenshot",
                   filename = "webassotest",
                   download = TRUE),
  downloadButton("download_plot",
                 label = " Download diagram (.pdf)"),
  downloadButton("download_results",
                 label = " Download results (.csv)")
)

# Vector PDF of the model-comparison diagram for publications.
output$download_plot <- downloadHandler(
  filename = function() paste0("webassotest_diagram_", Sys.Date(), ".pdf"),
  content  = function(file) {
    d <- ca_co()
    req(check_data_validity(d$ca, d$co)$valid)
    cairo_pdf(file, width = 11, height = 9)   # cairo keeps the \u2265 glyph
    on.exit(dev.off())
    gtm(d$ca, d$co, firth = input$use_firth, flev = flev_labels())
  }
)

# CSV download handler — one file, stacked sections with a shared set of
# columns (cells not relevant to a section are NA), so the file is a
# complete record of the analysis:
#
#   Section = "Settings":       app version, date, regression method,
#                               HWE test and genotype labels (Value column).
#   Section = "Input":          the entered genotype counts.
#   Section = "Model":          one row per OR contrast (Genotype has two),
#                               with Wald/Fisher CI and model AIC.
#   Section = "LRT":            every likelihood-ratio test on the diagram.
#   Section = "Allelic tests":  Cochran-Armitage trend and allelic test.
#   Section = "HWE":            Cases, Controls, Both; selected HWE method.
#
# gtm() is called for its invisible return value; the plot it draws is
# discarded since we are inside a download handler, not a render context.
output$download_results <- downloadHandler(
  filename = function() paste0("webassotest_results_", Sys.Date(), ".csv"),
  content  = function(file) {
    d     <- ca_co()
    ca    <- d$ca
    co    <- d$co
    req(check_data_validity(ca, co)$valid)
    hwe <- hwe_pvalues(ca, co, method = input$hwe_method)
    png(tempfile())   # discard the plot gtm() draws
    res <- gtm(ca, co, firth = input$use_firth, flev = flev_labels())
    dev.off()
    al  <- allelic_tests(ca, co)
    sections <- list(
      data.frame(Section  = "Settings",
                 Contrast = c("App version", "Date", "Regression", "HWE test", "Genotype labels"),
                 Value    = c(app_version, format(Sys.Date()),
                              if (input$use_firth) "Firth's penalized" else "Standard logistic",
                              if (input$hwe_method == "exact") "Exact" else "Chi-square",
                              paste(flev_labels(), collapse = "/"))),
      data.frame(Section = "Input", Contrast = c("Cases", "Controls"),
                 WT = c(ca[1], co[1]), Het = c(ca[2], co[2]), Hom = c(ca[3], co[3])),
      cbind(Section = "Model", res$models),
      cbind(Section = "LRT", setNames(res$tests, c("Contrast", "df", "Chisq", "p_value"))),
      cbind(Section = "Allelic tests", setNames(al, c("Contrast", names(al)[-1]))),
      data.frame(Section  = "HWE",
                 Contrast = c("Cases", "Controls", "Both"),
                 Chisq    = c(hwe$t.ca, hwe$t.co, hwe$t.2),
                 p_value  = c(hwe$p.ca, hwe$p.co, hwe$p.2))
    )
    cols <- unique(unlist(lapply(sections, names)))
    out  <- do.call(rbind, lapply(sections, function(x) {
      x[setdiff(cols, names(x))] <- NA
      x[cols]
    }))
    write.csv(out, file, row.names = FALSE)
  }
)
```


Row {data-height=1100}
-------------------------------------

### Result `r textOutput("mode_label", inline = TRUE)`

```{r}
# Parenthetical next to the "Result" heading showing which regression
# method produced the plot below (toggled via the use_firth checkbox).
output$mode_label <- renderText({
  paste0("(", if (input$use_firth) "Firth's penalized regression" else "Standard logistic regression", ")")
})

# plotOutput and renderPlot are declared separately (rather than using the
# inline renderPlot({}) shorthand) so that height can be set via CSS ("75vh")
# rather than being polled by JavaScript. The inline shorthand triggers a
# Shiny resize-observer loop that causes the plot to jump when other reactive
# outputs update; the explicit split avoids this.
#
# res = 96 pins rendering resolution to standard screen DPI for consistency.
# validate() shows a user-facing message instead of an R error when input fails
# check_data_validity().
plotOutput("result_plot", height = "75vh")

# model_summary holds gtm()'s invisible return (Model/OR/CI/Chisq/p/AIC data
# frame) so the AIC table below can reuse it without re-fitting the models.
model_summary <- reactiveVal(NULL)

output$result_plot <- renderPlot({
  d     <- ca_co()
  ca    <- d$ca
  co    <- d$co
  check <- check_data_validity(ca, co)
  validate(need(check$valid, check$msg))
  model_summary(gtm(ca, co, firth = input$use_firth, flev = flev_labels()))
}, res = 96)

# Compact AIC comparison table — lower AIC indicates better relative fit —
# shown beside the trend/allelic tests. Built with raw tags (like the HWE
# table above) rather than renderTable so the lowest-AIC row can be
# highlighted.
uiOutput("result_tables")
output$result_tables <- renderUI({
  req(model_summary())
  df   <- unique(model_summary()$models[, c("Model", "AIC")])
  best <- which.min(df$AIC)   # ignores NA automatically; first row wins on ties
  aic_rows <- lapply(seq_len(nrow(df)), function(i) {
    tags$tr(
      style = if (i == best) "font-weight:bold; background-color:#c8e6c9;" else NULL,
      tags$td(df$Model[i]),
      tags$td(formatC(df$AIC[i], format = "f", digits = 1))
    )
  })
  d  <- ca_co()
  al <- allelic_tests(d$ca, d$co)
  al_rows <- lapply(seq_len(nrow(al)), function(i) tags$tr(
    tags$td(al$Test[i]),
    tags$td(formatC(al$Chisq[i], format = "f", digits = 3)),
    tags$td(fmt_p(al$p_value[i])),
    tags$td(if (is.na(al$OR[i])) "\u2014" else fmt_ci(unlist(al[i, c("OR", "CI_lower", "CI_upper")])))
  ))
  tags$div(style = "display:flex; flex-wrap:wrap; gap:24px; align-items:flex-start;",
    tags$table(
      class = "table table-sm table-bordered", style = "width:300px;",
      tags$thead(tags$tr(tags$th("Model"), tags$th("AIC"))),
      tags$tbody(aic_rows)),
    tags$table(
      class = "table table-sm table-bordered", style = "width:auto;",
      tags$thead(tags$tr(tags$th("Trend tests"), tags$th("\u03c7\u00b2"),
                         tags$th("p"), tags$th("Allelic OR (95% CI)"))),
      tags$tbody(al_rows))
  )
})
```


About {#about}
==========================

Column {data-width=600}
-----------------------------------------------------------------------

### About Web-Assotest

Web-Assotest is a browser-based tool for analysing the association between a genetic variant and a binary outcome (cases vs. controls) using a 3×2 genotype count table. It fits four standard genetic models — **dominant**, **recessive**, **additive** (log-additive: the same OR per allele copy), and **genotype** — and displays odds ratios with 95% confidence intervals and likelihood-ratio test statistics for each, allowing the user to assess which models are consistent with the data.

The tool also tests for departure from **Hardy-Weinberg equilibrium (HWE)** in cases, controls, and both combined, using either a chi-square test or the exact test of Wigginton, Cutler & Abecasis (2005).

For an overview and explanation of the models, see the [course slides](https://ekstroem.com/download/assotestslides.pdf) on case-control studies and genetic association analysis. Web-Assotest is [widely used in scientific publications](http://scholar.google.com/scholar?q=web-assotest).

Web-Assotest uses [R](http://www.r-project.org/) for all computations. See the [R Primer book](http://www.rprimer.dk/) for an introduction to statistical analysis in R.

Two 1-df tests are also reported: the **Cochran-Armitage trend test** and the **allelic test** (allele counts in a 2×2 table, with the allelic OR). The allelic test assumes HWE; the trend test does not. **Flip allele coding** swaps which homozygote is the reference, which swaps the meaning of the dominant and recessive models.

**Note:** cells with zero counts are handled using Fisher's exact test for the affected contrasts. Take care when many cells contain zeros.

### References

Wigginton JE, Cutler DJ, Abecasis GR (2005). A Note on Exact Tests of Hardy-Weinberg Equilibrium. *American Journal of Human Genetics*, 76(5):887–893.
